---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This assignment analyzes data from a personal activity monitoring device.  
  
The variables included in this dataset are:  
- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)  
- date: The date on which the measurement was taken in YYYY-MM-DD format  
- interval: Identifier for the 5-minute interval in which measurement was taken  


```{r working directory, echo=FALSE}
setwd("C:/Users/smythto/Documents/_Never Backup/Dropbox/Coursera/Reproducible Research/Project 1")
```
##Loading and preprocessing the data  
Read the data from file:
```{r load data}
activityRaw <- read.csv("activity.csv", header = TRUE, na.strings = "NA")
```

Raw data dimensions and 10 sample records:
```{r overview}
dim(activityRaw)
activityRaw[sample(nrow(activityRaw),10),]  

```

NA values for field "steps"" will be exluded from analysis as they represent a lack of a measurement (not necessarily the same a an observation of "0"):
```{r na}

activity <- subset(activityRaw, is.na(steps) == FALSE)   
activityNa <- subset(activityRaw,  is.na(steps) == TRUE)
numNaRecs <- nrow(activityNa)  
`````

Cleansed data dimensions:
```{r overview of cleansed}
dim(activity)
```

Aggregate clean data to get mean and sum of steps per day:
```{r aggregate}
library("plyr")
sumByDay<- ddply(activity,.(date),summarize, sumSteps = sum(steps))
```
##What is mean total number of steps taken per day?  
```{r mean_of_steps}
hist(sumByDay$sumSteps, main = "Frequency of total steps taken per day", xlab ="Total steps taken per day", ylab = "Number of days")  
stepsMedian <- round(median(sumByDay$sumStep), 1)
stepsMean <- round(mean(sumByDay$sumStep),1)
abline(v=stepsMedian,lty = 2, lwd=2,col = "red")  
abline(v=stepsMean,lty = 3,lwd=1,col = "blue")  
legend("topright",c(paste0("median (", round(stepsMedian,1) ,")" ) ,paste0("mean (" , round (stepsMean,1) , ")")),lwd=2,col = c("red","blue"))
```

Aggregate clean data to get average steps per interval:  
```{r aggregate for intervals}
avgByInt<- ddply(activity,.(interval),summarize, avgSteps = mean(steps))
```

##What is the average daily activity pattern?  
``` {r plot_daily}

plot(avgByInt$interval, avgByInt$avgSteps, type = "l", main = "Average Daily Activity Pattern", xlab ="5 minute time interval", ylab = "Average steps taken (across all days)", col = "blue")
maxInt <- avgByInt[avgByInt$avgSteps == max(avgByInt$avgSteps),c("interval")]
maxSteps <-round(max(avgByInt$avgSteps),2)
```
**Interval with average maximum steps**  
The maxinum number of steps (`r maxSteps`) occurred in interval `r maxInt`.  

##Imputing missing values     
As noted above, NA values were observed in the data.  `r numNaRecs` records were observed with missing values in the original activity monitoring data.  

**Implement impute strategy, create new data set with imputed steps**   
```{r impute}
library("mice")
imputedData <- mice(activityRaw,m=5, maxit = 5, method = "pmm", seed = 50)
##summary(imputedData)
activityImputed <-  complete(imputedData, 2)

sumByDayImputed<- ddply(activityImputed,.(date),summarize, sumSteps = sum(steps))
```

**Imputed data histogram**  
``` {r impute_histogram}

hist(sumByDayImputed$sumSteps, main = "Frequency of total steps taken/day (with imputed missing values for steps)", xlab ="Total steps taken per day", ylab = "Number of days")  
stepsMedianImputed <- round(median(sumByDayImputed$sumStep),1)
stepsMeanImputed <- round(mean(sumByDayImputed$sumStep),1)
abline(v=stepsMedianImputed,lty = 2, lwd=2,col = "red")  
abline(v=stepsMeanImputed,lty = 3,lwd=1,col = "blue")  
legend("topright",c(paste0("median (", round(stepsMedianImputed,1) ,")" ) ,paste0("mean (" , round (stepsMeanImputed,1) , ")")),lwd=2,col = c("red","blue"))
```


**Impact of imputing data**   
Comparing this chart with imputed steps to the chart above where missing steps were removed (in requirement 3), it is observed that the mean changed from `r format(stepsMean, digits = 1)` to `r format(stepsMeanImputed, digits = 1)` and the median change from `r format(stepsMedian, digits = 1)` to `r format(stepsMedianImputed, digits = 1)`.

##Are there differences in activity patterns between weekdays and weekends?  
```{r Weekday_analysis}
activityImputed$date = as.Date(activityImputed$date)
activityImputed$DayOfWeek <- ifelse(weekdays(activityImputed$date) == "Saturday" | weekdays(activityImputed$date) == "Sunday", "Weekend","Weekday")

avgByIntWD<- ddply(activityImputed,.(interval, DayOfWeek),summarize, avgSteps = mean(steps))

library("lattice")
xyplot(avgSteps ~interval | DayOfWeek, data = avgByIntWD, type = "l", layout=c(1,2), xlab = "Interval", ylab = "Number of Steps", main = "Average Daily Activity Pattern")
```

